\chapter*{Abstract}\label{chp:abstract}

%% Abstract page

%% The third page must have a short reference (abstract).

%% The reference must:
%% Be written in English.
%% Contain the name of the dissertation, possible subtitle, name of the author, department and 'Chalmers University of Technology'.
%% If the thesis is written in Swedish this should also be stated.
%% Be concise and do not exceed more than approximately 250 words.
%% Provide a brief, easily understood overview of the essential content of the thesis (problems, methods, results).
%% Conclude with a maximum of ten key words of significance for computerised information systems.

%% The reference can also be printed on the back of the presentation sheet, see below.

% \textbf{\lictitle}\\
% \textit{\licsubtitle}\\
% \textsc{\licauthor}\\
% \licdepartment\\
% \textsc{\licuniversity}\\

Constraint Grammar (CG) is a robust and language-independent formalism 
for part-of-speech tagging and shallow parsing. 
A grammar consists of disambiguation rules for initially ambiguous, 
morphologically analysed text: the correct analysis for the sentence 
is attained by removing improper readings from ambiguous words.
Wide-coverage Constraint Grammars, containing some thousands of rules, 
generally achieve very high accuracy for free text: 
thanks to the flexible formalism, new rules can be written 
to address even the rarest phenomena, without compromising the general tendencies.
Even with a smaller set of rules, CG-based taggers have been 
competetive with statistical taggers.
In addition, CG has been used in more experimental settings: dependency syntax,
dialogue system for language learning %\cite{antonsen2009dialogue}
and anaphora resolution. %\cite{puolakainen2015anaphora}.

This thesis presents two contributions to the study of CG.
Firstly, we model a parallel variant of CG as a 
Boolean satisfiability (SAT) problem, and
implement a CG-engine by using a SAT-solver.
This is attractive for several reasons: formal logic is well-studied,
and serves as an abstract language to reason about the properties of
CG. Using SAT-solver as a backend, we can exploit decades of 
research into optimising SAT-solvers.
In addition, we suggest two novel alternatives for 
conflict handling in the parallel scheme.
%: this brings performance gains compared to previous implementations of parallel CG.

Secondly, we implement a SAT-based tool for analysing sequential CGs.
Sequential CG is the most commonly used variant: rules are applied in order, 
and actions are irreversible.
We apply \formulation{a common technique from software verification}, 
and model the state of all possible sentences after applying each rule;
this enables grammar writers to see whether some rule or rule set prevents
others from applying, or request example sequences that trigger a given rule.

\bigskip
\noindent
\textbf{Keywords}: \emph{\lickeywords}
