\chapter{Background}

In this chapter, we present the two components of this thesis: Constraint Grammar and SAT.
Section~\ref{sec:cg-intro} broadly introduces the CG formalism and some of the different variants, along with the historical context and related work.
For a description of any specific CG implementation, we direct the reader to the original source: CG-1 \cite{karlsson1990cgp,karlsson1995constraint}, CG-2 \cite{tapanainen1996}, VISL CG-3 \cite{bick2015,vislcg3}.
Section~\ref{sec:SAT-intro} is a brief and high-level introduction to SAT, aimed to a reader with no prior knowledge on SAT. For a more thorough description, we recommend \cite{biere2009handbook}.

\input{chapters/cg-intro}

\input{chapters/sat-intro}


\section{Summary}

In this section, we have presented the theoretical background used in this thesis. 
We have introduced Constraint Grammar: a formalism for writing disambiguation rules; and Boolean satisfiability: methods for solving whether a given propositional formula can evaluate to true.
In the following chapters, we will connect the two branches. First, we look into existing research in connecting CG to logic, and formulate a SAT-instance for two varieties of CG. After establishing a SAT-encoding, we use it, much in the spirit of software verification, to show interesting properties of a given grammar. 


%highlighting some key properties and relating it to other grammar formalisms.
% CG is relatively young, popularised 20 years ago, and throughout its
% history, it has been in favour of linguists more than computer
% scientists, which might explain why its formal background has not been
% studied extensively---notable exceptions being \cite{lager98}, \cite{tapanainen1999phd} and \cite{lager_nivre01}.

 


% \todo{As for its descriptive power, we can again compare CG to CFG. If
% badly constructed, both CG and CFG can analyse almost everything,
% making it useless for language description. If we accept that a
% possibility to construct a bad grammar doesn't revoke CFG's status as
% a formalism, it shouldn't do so for CG either. One can always compare
% the relative ease of constructing a good grammar between formalisms,
% but that is another topic already.}

% As a bottom line, CG provides a lightweight framework where already a
% moderate amount of rules can be useful (e.g. \cite{lene_trond2011} with 115
% rules), although high quality requires in general thousands of rules.
% Like any formalism, CG itself leaves the users with much freedom; a
% single grammar can be anything from completely ungrammatical rules to near-perfect description of language.



